---
title: "RShiny Lab: Part II"
subtitle: "Workshop on Plotting in R"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
---

```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot")
)
```

```{r,echo=FALSE,child="assets/header-lab.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r,echo=FALSE,results='hide',warning=FALSE,message=FALSE}
library(shiny)
library(ggplot2)
```


# Updating widgets

Widgets can be updated with new values dynamically. `observe()` and `observeEvent()` functions can monitor the values of interest and update relevant widgets.

```{r}
shinyApp(
ui=fluidPage(
  selectInput("data_input",label="Select data",choices=c("mtcars","faithful","iris")),
  selectInput("header_input",label="Select column name",choices=NULL),
  plotOutput("plot_output",width="400px")
),
server=function(input,output,session) {
  getdata <- reactive({ get(input$data_input, 'package:datasets') })
  
  observe({
    updateSelectInput(session,"header_input",label="Select column name",choices=colnames(getdata()))
  })
  
  output$plot_output <- renderPlot({
    #shiny::req(input$header_input)
    #validate(need(input$header_input %in% colnames(getdata()),message="Incorrect column name."))
    hist(getdata()[, input$header_input],xlab=input$header_input,main=input$data_input)
  })
},
options=list(height=600))
```

In this example, the user selects a dataset and a column from the selected dataset to be plotted as a histogram. The column name selection widget must automatically update it's choices depending on the selected dataset. This achieved using `observe()` where the `updateSelectInput()` function updates the selection choices. Notice that a third option `session` is in use in the server function. ie; `server=function(input,output,session)`. And `session` is also the first argument in `updateSelectInput()`. Session keeps track of values in the current session.

When changing the datasets, we can see that there is a short red error message. This is because, after we have selected a new dataset, the old column name from the previous dataset is searched for in the new dataset. This occurs for a short time and causes the error. This can be fixed using careful error handling. We will discuss this in another section.